rgbif: R client for working with GBIF species occurrence data

### 2015-10-05 manuscript.Rmd
more writing ✏ ✏

---
#layout: preprint
layout: review, 11pt
linenumbers: true
title: "rgbif: a package for working with species occurrence data in R"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      University of California, Berkeley, CA, USA
abstract: |
      1. xxx

@@

bibliography: components/references.bib
csl: components/methods-in-ecology-and-evolution.csl
documentclass: components/elsarticle

@@

The download API syntax is similar to the occurrence search API
in that the same parameters are used, but the way in which the query
is defined is different. For example, in the download API you can
do greater than searches, whereas you can not do that in the
occurrence search API. Thus, we can't make the query interace exactly
the same for both search and download functions.

Using the download service can be as few as three steps.

Request data download given a query. Here, we xxxx

```{r}
"xxx"
```

@@

```{r}
(res <- occ_download_get("0000066-140928181241064", overwrite = TRUE))
```

@@

@@

The search API is very similar to the download API, but is meant for smaller
data acquisition jobs.

@@

All software, scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI corresponds to a snapshot of the GitHub repository at [github.com/sckott/msrgbif](https://github.com/sckott/msrgbif).

@@

[mit]: http://choosealicense.com/licenses/mit/
[ghrgbif]: https://github.com/ropensci/rgbif
[cranrgbif]: https://cran.rstudio.com/web/packages/rgbif/
[gbifdocs]: http://www.gbif.org/developer/summary

>>>>

---
layout: review, 11pt
linenumbers: true
title: "rgbif: R client for working with GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      University of California, Berkeley, CA, USA
abstract: |
      1. xxx

@@

bibliography: components/references.bib
csl: components/peerj.csl
documentclass: components/elsarticle

@@

The download API syntax is similar to the occurrence search API
in that the same parameters are used, but the way in which the query
is defined is different. For example, in the download API you can
do greater than searches (i.e., `latitude > 50`), whereas you can not
do that in the occurrence search API. Thus, we can't make the query
interace exactly the same for both search and download functions.

Using the download service can be as few as three steps: 1) Request data
via a search; 2) Download data; 3) Import data into R.

Request data download given a query. Here, we search for the taxon key
`3119195`, which is the key for _Helianthus annuus_
(http://www.gbif.org/species/3119195).

```{r}
occ_download('taxonKey = 3119195')
```

@@

```{r}
(res <- occ_download_get("0000840-150615163101818", overwrite = TRUE))
```

@@

#### Downloaded data format

The downloaded dataset from GBIF is actually a Darwin Core Archive (DwC-A), an
internationally recognized biodiversity informatics standard
(http://rs.tdwg.org/dwc/). The DwC-A downloaded is a compressed folder with
a number of files, including metadata, citations for each of the datasets included
in the download, and the data itself, in separate files for each dataset as well
as one single `.txt` file. In `occ_download_import()`, we simply fetch data from the
`.txt` file. If you want to dig into the metadata, citations, etc., it is easily
accessible from the folder on your computer.

@@

The search API follows the GBIF API and is broken down into the following functions:

* `occ_count()`
* `occ_search()`
* `occ_get()`
* `occ_metadata()`

The main search work-horse is `occ_search()`. This function allows very flexible
search definitions. In addition, this function does paging internally, making it
such that the user does not have worry about the 300 records per request limit -
but of course we can't go over the 200,000 maximum limit.

...

#### Cleaning data

GBIF provides optional data issues with each occurrence record. These issues fall
into many different pre-defined classes, covering issues with taxonomic names,
geographic data, and more (see `occ_issues_lookup()` to find out more information
on GBIF issues; and the same data on [GBIF's development site][gbifissues]).

`occ_issues()` provides a way to easily filter data downloaded via `occ_search()`
based on GBIF issues.

```{r}
out <- occ_search(issue='DEPTH_UNLIKELY', limit = 500)
NROW(out)
out %>% occ_issues(-cudc) %>% .$data %>% NROW
```

## Use cases

...

@@

All scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI corresponds to a snapshot of the GitHub repository at [github.com/sckott/msrgbif](https://github.com/sckott/msrgbif). Software
can be found at [github.com/ropensci/rgbif](https://github.com/ropensci/rgbif), under the
open and permissive MIT license.

@@

[mit]: http://choosealicense.com/licenses/mit/
[ghrgbif]: https://github.com/ropensci/rgbif
[cranrgbif]: https://cran.rstudio.com/web/packages/rgbif/
[gbifdocs]: http://www.gbif.org/developer/summary
[gbifissues]: http://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceIssue.html

### 2015-10-05 manuscript.tex
more writing ✏ ✏

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={rgbif: a package for working with species occurrence data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{rgbif: a package for working with species occurrence data in R}
    \author[cstar]{Scott Chamberlain\corref{c1}}
   \ead{scott(at)ropensci.org} 
   \cortext[c1]{Corresponding author}
      \address[cstar]{University of California, Berkeley, CA, USA}

@@

Perhaps the most fundamental element in many fields of ecology is the
inividual. How many individuals of each species in a given location
forms the basis for many sub-fields of ecology and evolution. Some
research questions necessitate collecting new data, while others can
easily take advantage of existing data. In fact, some ecology fields are
built largely on existing data, e.g., macro-ecology (Brown 1995; Beck
\emph{et al.} 2012).

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. In addition, the
pool of questions we can answer becomes much larger with more and better
data. In addition to wide utility, this data is important for
conservation. Biodiversity loss is one of the greatest challenges of our
time (Pimm \emph{et al.} 2014). Some have called this the sixth great
mass extinction (Ceballos \emph{et al.} 2015). Given this challenge
there is a great need for data on specimen records, whether collected
from live sightings in the field or specimens in museums.

@@

Herein, we describe a library (rgbif (Chamberlain \emph{et al.})) for
working with GBIF data in the R programming environment (R Core Team
2014). R is an extremely widely used language in academia, and in
non-profit and private sectors. Importantly, R makes it easy to do all
of the steps of the research process, including data management, data
manipulation and cleaning, statistics, and vizualization. Thus, an R
client for getting GBIF data is a powerful tool for reproducible
research.

@@

The download API syntax is similar to the occurrence search API in that
the same parameters are used, but the way in which the query is defined
is different. For example, in the download API you can do greater than
searches, whereas you can not do that in the occurrence search API.
Thus, we can't make the query interace exactly the same for both search
and download functions.

Using the download service can be as few as three steps.

Request data download given a query. Here, we xxxx

\begin{Shaded}
\begin{Highlighting}[]
\StringTok{"xxx"}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] "xxx"
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{(res <-}\StringTok{ }\KeywordTok{occ_download_get}\NormalTok{(}\StringTok{"0000066-140928181241064"}\NormalTok{, }\DataTypeTok{overwrite =} \OtherTok{TRUE}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
<<gbif downloaded get>>
  Path: ./0000066-140928181241064.zip
  File size: 0.14 MB
\end{verbatim}

@@

@@

The search API is very similar to the download API, but is meant for
smaller data acquisition jobs.

@@

All software, scripts and data used in this paper can be found in the
permanent data archive Zenodo under the digital object identifier (DOI).
This DOI corresponds to a snapshot of the GitHub repository at
\href{https://github.com/sckott/msrgbif}{github.com/sckott/msrgbif}.

@@

Beck, J., Ballesteros-Mejia, L., Buchmann, C.M., Dengler, J., Fritz,
S.A., Gruber, B., Hof, C., Jansen, F., Knapp, S., Kreft, H., Schneider,
A.-K., Winter, M. \& Dormann, C.F. (2012). Whats on the horizon for
macroecology? \emph{Ecography}, \textbf{35}, 673--683. Retrieved from
\url{http://dx.doi.org/10.1111/j.1600-0587.2012.07364.x}

Brown, J.H. (1995). \emph{Macroecology}. University of Chicago Press.

Ceballos, G., Ehrlich, P.R., Barnosky, A.D., Garcia, A., Pringle, R.M.
\& Palmer, T.M. (2015). Accelerated modern human-induced species losses:
Entering the sixth mass extinction. \emph{Science Advances}, \textbf{1},
e1400253--e1400253. Retrieved from
\url{http://dx.doi.org/10.1126/sciadv.1400253}

Chamberlain, S., Ram, K., Barve, V. \& Mcglinn, D. \emph{Rgbif:
Interface to the global 'biodiversity' information facility 'aPI'}.
Retrieved from \url{https://github.com/ropensci/rgbif}

Pimm, S.L., Jenkins, C.N., Abell, R., Brooks, T.M., Gittleman, J.L.,
Joppa, L.N., Raven, P.H., Roberts, C.M. \& Sexton, J.O. (2014). The
biodiversity of species and their rates of extinction, distribution, and
protection. \emph{Science}, \textbf{344}, 1246752--1246752. Retrieved
from \url{http://dx.doi.org/10.1126/science.1246752}

R Core Team. (2014). \emph{R: A language and environment for statistical
computing}. R Foundation for Statistical Computing, Vienna, Austria.
Retrieved from \url{http://www.R-project.org/}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={rgbif: R client for working with GBIF species occurrence data},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{rgbif: R client for working with GBIF species occurrence data}
    \author[cstar]{Scott Chamberlain\corref{c1}}
   \ead{scott(at)ropensci.org} 
   \cortext[c1]{Corresponding author}
      \address[cstar]{University of California, Berkeley, CA, USA}

@@

Perhaps the most fundamental element in many fields of ecology is the
inividual. How many individuals of each species in a given location
forms the basis for many sub-fields of ecology and evolution. Some
research questions necessitate collecting new data, while others can
easily take advantage of existing data. In fact, some ecology fields are
built largely on existing data, e.g., macro-ecology (Brown, 1995; Beck
et al., 2012).

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. In addition, the
pool of questions we can answer becomes much larger with more and better
data. In addition to wide utility, this data is important for
conservation. Biodiversity loss is one of the greatest challenges of our
time (Pimm et al., 2014). Some have called this the sixth great mass
extinction (Ceballos et al., 2015). Given this challenge there is a
great need for data on specimen records, whether collected from live
sightings in the field or specimens in museums.

@@

Herein, we describe a library (rgbif (Chamberlain et al.)) for working
with GBIF data in the R programming environment (R Core Team, 2014). R
is an extremely widely used language in academia, and in non-profit and
private sectors. Importantly, R makes it easy to do all of the steps of
the research process, including data management, data manipulation and
cleaning, statistics, and vizualization. Thus, an R client for getting
GBIF data is a powerful tool for reproducible research.

@@

The download API syntax is similar to the occurrence search API in that
the same parameters are used, but the way in which the query is defined
is different. For example, in the download API you can do greater than
searches (i.e., \texttt{latitude \textgreater{} 50}), whereas you can
not do that in the occurrence search API. Thus, we can't make the query
interace exactly the same for both search and download functions.

Using the download service can be as few as three steps: 1) Request data
via a search; 2) Download data; 3) Import data into R.

Request data download given a query. Here, we search for the taxon key
\texttt{3119195}, which is the key for \emph{Helianthus annuus}
(\url{http://www.gbif.org/species/3119195}).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{occ_download}\NormalTok{(}\StringTok{'taxonKey = 3119195'}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
<<gbif download>>
  Username: sckott
  E-mail: myrmecocystus@gmail.com
  Download key: 0004098-150922153815467
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{(res <-}\StringTok{ }\KeywordTok{occ_download_get}\NormalTok{(}\StringTok{"0000840-150615163101818"}\NormalTok{, }\DataTypeTok{overwrite =} \OtherTok{TRUE}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
<<gbif downloaded get>>
  Path: ./0000840-150615163101818.zip
  File size: 3.19 MB
\end{verbatim}

@@

\paragraph{Downloaded data format}\label{downloaded-data-format}

The downloaded dataset from GBIF is actually a Darwin Core Archive
(DwC-A), an internationally recognized biodiversity informatics standard
(\url{http://rs.tdwg.org/dwc/}). The DwC-A downloaded is a compressed
folder with a number of files, including metadata, citations for each of
the datasets included in the download, and the data itself, in separate
files for each dataset as well as one single \texttt{.txt} file. In
\texttt{occ\_download\_import()}, we simply fetch data from the
\texttt{.txt} file. If you want to dig into the metadata, citations,
etc., it is easily accessible from the folder on your computer.

@@

The search API follows the GBIF API and is broken down into the
following functions:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \texttt{occ\_count()}
\item
  \texttt{occ\_search()}
\item
  \texttt{occ\_get()}
\item
  \texttt{occ\_metadata()}
\end{itemize}

The main search work-horse is \texttt{occ\_search()}. This function
allows very flexible search definitions. In addition, this function does
paging internally, making it such that the user does not have worry
about the 300 records per request limit - but of course we can't go over
the 200,000 maximum limit.

\ldots{}

\paragraph{Cleaning data}\label{cleaning-data}

GBIF provides optional data issues with each occurrence record. These
issues fall into many different pre-defined classes, covering issues
with taxonomic names, geographic data, and more (see
\texttt{occ\_issues\_lookup()} to find out more information on GBIF
issues; and the same data on
\href{http://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceIssue.html}{GBIF's
development site}).

\texttt{occ\_issues()} provides a way to easily filter data downloaded
via \texttt{occ\_search()} based on GBIF issues.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{out <-}\StringTok{ }\KeywordTok{occ_search}\NormalTok{(}\DataTypeTok{issue=}\StringTok{'DEPTH_UNLIKELY'}\NormalTok{, }\DataTypeTok{limit =} \DecValTok{500}\NormalTok{)}
\KeywordTok{NROW}\NormalTok{(out)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] 4
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{out %>%}\StringTok{ }\KeywordTok{occ_issues}\NormalTok{(-cudc) %>%}\StringTok{ }\NormalTok{.$data %>%}\StringTok{ }\NormalTok{NROW}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] 2
\end{verbatim}

\subsection{Use cases}\label{use-cases}

\ldots{}

@@

All scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI
corresponds to a snapshot of the GitHub repository at
\href{https://github.com/sckott/msrgbif}{github.com/sckott/msrgbif}.
Software can be found at
\href{https://github.com/ropensci/rgbif}{github.com/ropensci/rgbif},
under the open and permissive MIT license.

@@

Beck J., Ballesteros-Mejia L., Buchmann CM., Dengler J., Fritz SA.,
Gruber B., Hof C., Jansen F., Knapp S., Kreft H., Schneider A-K., Winter
M., Dormann CF. 2012. Whats on the horizon for macroecology?
\emph{Ecography} 35:673--683.

Brown JH. 1995. \emph{Macroecology}. University of Chicago Press.

Ceballos G., Ehrlich PR., Barnosky AD., Garcia A., Pringle RM., Palmer
TM. 2015. Accelerated modern human-induced species losses: Entering the
sixth mass extinction. \emph{Science Advances} 1:e1400253--e1400253.

Chamberlain S., Ram K., Barve V., Mcglinn D. \emph{Rgbif: Interface to
the global 'biodiversity' information facility 'aPI'}.

Pimm SL., Jenkins CN., Abell R., Brooks TM., Gittleman JL., Joppa LN.,
Raven PH., Roberts CM., Sexton JO. 2014. The biodiversity of species and
their rates of extinction, distribution, and protection. \emph{Science}
344:1246752--1246752.

R Core Team. 2014. \emph{R: A language and environment for statistical
computing}. Vienna, Austria: R Foundation for Statistical Computing.
